---
title: "Qploidy2 QuickStart Template"
subtitle: "Copy number estimation from standardized data"
author: "Breeding Insight"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    css: BIcolors_template.css
    includes:
      in_header: header.html
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    code_folding: show
    df_print: paged
params:
  standardization_file: "standardization.tsv.gz"
  output_prefix: "qploidy_results"
  n_cores: 2
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 10,
  out.width = "40%"
)
```

# Parameters

Define key parameters before running the analysis. Adjust these to match your dataset.

```{r, eval=FALSE, echo=FALSE}
params <- list()

params$standardization_file <- "standardization.tsv.gz"
params$n_cores <- 3
params$output_prefix <- "alfalfa_test"
```

```{r parameters}
# Path to your standardized data file
standardization_file <- params$standardization_file

# Index of the example sample used throughout this template for single-sample steps.
# Increase if samples[3] is not informative (e.g., low heterozygosity).
example_sample_idx <- 3

# Number of CPU cores for parallel computation
n_cores <- params$n_cores

# Copy number range to test (adjust based on your expected ploidy)
cn_grid <- c(2, 3, 4, 5, 6)

# Predominant ploidy for re-standardization (should match the most frequent ploidy in your dataset)
ploidy_standardization <- 4

# Quality thresholds for filter_hmm_CN
min_CN_reliability <- 0.25
min_post_max <- 0.6

# Output file prefix for exported results
output_prefix <- params$output_prefix
```

# Introduction

This template demonstrates the Qploidy2 workflow starting from a pre-computed `standardization.tsv.gz`
file. It covers quality assessment, BAF model selection, Hidden Markov Model (HMM)-based copy number
estimation, visualization, interactive curation, and VCF export.

If you have not yet produced the standardization file, follow the full tutorial
(`Qploidy_alfalfa_tutorial.Rmd`) for input preparation and standardization steps.

# Load Standardized Data

```{r}
library(Qploidy2)
packageVersion("Qploidy2")

pkg_ver    <- as.character(packageVersion("Qploidy2"))
output_dir <- paste0(output_prefix, "_v", pkg_ver)
images_dir <- file.path(output_dir, "images")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(images_dir, showWarnings = FALSE, recursive = TRUE)

qploidy_standardization <- read_qploidy_standardization(standardization_file)
samples <- unique(qploidy_standardization$data$SampleName)
sample_id <- samples[example_sample_idx]

cat("Loaded", length(samples), "samples\n")
cat("Example sample:", sample_id, "\n")
```

# Quality Assessment and Visualization

Visualize the effect of standardization for a single sample and explore BAF and Z-score distributions:

```{r}
plot_qploidy_standardization(x = qploidy_standardization,
                             sample = sample_id,
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))

plot_qploidy_standardization(x = qploidy_standardization,
                             sample = sample_id,
                             type = c("zscore", "R"))
```

See `?plot_qploidy_standardization` for all available plot types.


# HMM Copy Number Estimation (Single Sample)

Estimate copy number for a sample using a Hidden Markov Model. This step integrates standardized BAF
and Z-score information across genomic regions to infer CNV segments.
Find all parameter descriptions in `?hmm_estimate_CN`.

**Warning**: Avoid testing copy numbers that are not biologically plausible for your species to reduce
the number of evaluated scenarios and improve runtime.

```{r}
res <- hmm_estimate_CN(qploidy_standarize_result = qploidy_standardization,
                       sample_id = sample_id,
                       cn_grid = cn_grid)

res
```

Note that the HMM multipoint approach considers marker position order based on the reference genome
used. Therefore, changing the reference genome may change the results.

`Qploidy2` offers several ways to visualize the estimated karyotype:

```{r}
plot_karyotype(res, sample_name = sample_id, nrow = 2)
```

Color by the CN call final likelihood (how confident the model was for each marker):

```{r}
plot_karyotype(res, sample_name = sample_id, nrow = 2, color_by = "post_max_CN")
```

Color by the BAF weight — windows inferred from both BAF and Z-score have weight 0.5; windows with no
heterozygous genotypes are inferred from Z-score only and have BAF weight close to 0:

```{r}
plot_karyotype(res, sample_name = sample_id, nrow = 2, color_by = "w_baf")
```

The `plot_cn_track` function provides a detailed diagnostic of the process:

```{r}
p <- plot_cn_track(hmm_CN = res,
                   qploidy_standarize_result = qploidy_standardization,
                   sample_id = sample_id,
                   show_window_lines = FALSE,
                   z_by_mean = TRUE)

p$arranged
```

The CN segments in the bottom track are coloured by `CN_reliability`, a joint confidence score
calculated as `post_max × normalized_w_baf`. It summarises both the HMM's certainty about the
called CN state and whether the call was supported by BAF evidence:

- **Range 0–1**: a score of `1` means the HMM assigned the highest posterior to the called CN
  state *and* the window contains enough heterozygous markers for BAF support to also be
  confident.
- **Low values near 0**: the call relies almost entirely on Z-score (total depth) because the
  window lacks heterozygous markers (`w_baf ≈ 0`). By default, such windows are capped at
  `0.25 × post_max`, so even a perfect HMM call cannot exceed 0.25 without BAF evidence.
- **Interpreting the plot**: bright segments indicate high-confidence calls backed by both
  signals; faded segments warrant manual inspection. Use `filter_hmm_CN(min_CN_reliability = ...)`
  in the [Filter by CN Call Quality Metrics](#filter-by-cn-call-quality-metrics) section to
  mask low-confidence windows before export.


# HMM Copy Number Estimation (Multiple Samples)

Run HMM for all samples in parallel. The BAF model is selected internally using
`select_best_baf_model`; define the grids (`bw`, `add_uniform_grid`, and `uniform_weight_grid`)
according to what you observed in the single-sample step above.

**Note**: Running HMM for many samples can be computationally intensive. Adjust `n_cores` based on
your available resources and dataset size.

```{r}
multi_esti <- hmm_estimate_CN_multi(qploidy_standarize_result = qploidy_standardization,
                                    sample_ids = "all",
                                    n_cores = n_cores,
                                    cn_grid = cn_grid)
```

Visualize a CNV summary for multiple samples:

```{r}
compare_cn_track(hmm_CN = multi_esti,
                 samples_to_plot = samples[1:min(30, length(samples))],
                 facet_ncol = 8)
```

# Filter by CN Call Quality Metrics

`filter_hmm_CN()` masks CN calls that do not meet user-defined quality thresholds by replacing
`CN_call`, `post_max`, `CN_reliability`, and all `post_CN*` posteriors with `NA`. Masked windows are
treated as missing by all downstream plotting and export functions.

The two most commonly used thresholds are:

* `min_CN_reliability` — joint confidence combining HMM posterior probability and BAF weight.
  A value of `0` means no BAF evidence at all; `1` means full BAF support with a confident HMM call.
  Windows supported only by Z-score cannot exceed `z_no_baf_scale × post_max` in reliability
  (default cap = `0.25 × post_max`).
* `min_post_max` — the raw HMM posterior probability for the called CN state, regardless of BAF
  support. Useful when you want to retain Z-score-only calls but still discard low-confidence HMM
  calls.

```{r}
multi_esti_filtered <- filter_hmm_CN(multi_esti,
                                     min_CN_reliability = min_CN_reliability,
                                     min_post_max = min_post_max)
```

# Summarized Plots and Tables

After filtering, get a high-level overview of the entire population.

`count_types()` classifies every sample as **euploid** (all chromosomes match the sample-level copy
number), **aneuploid** (one or more chromosomes carry a different but uniform CN — whole-chromosome
gain or loss), or **segmental** (at least one chromosome has windows with mixed CN, indicating a
partial chromosomal change):

```{r}
types_df <- count_types(multi_esti)
types_df
str(types_df)

types_df_filtered <- count_types(multi_esti_filtered)
types_df_filtered
str(types_df_filtered)
```

Use the returned lists to visualize samples of each type:

```{r}
# Adjust indices if the first sample of a type is not informative
if (length(types_df$aneuploid) > 0) {
  plot_karyotype(multi_esti,
                 sample_name = types_df$aneuploid[1], color_by = "CN_reliability")

  plot_karyotype(multi_esti,
                 sample_name = types_df$aneuploid[1], color_by = "CN_reliability", add_ghost = FALSE)
} else {
  message("No aneuploid samples found in multi_esti.")
}

if (length(types_df_filtered$segmental) > 0) {
  plot_karyotype(multi_esti_filtered,
                 sample_name = types_df_filtered$segmental[1], color_by = "CN_reliability")
} else {
  message("No segmental samples found in multi_esti_filtered.")
}
```

`compare_cn_track_summary()` produces a karyotype-style matrix with samples in rows and chromosomes
in columns. Samples are ordered by ascending sample-level CN and, within each CN block, by ploidy
category (euploid → aneuploid → segmental). Chromosomes with mixed CN are shown in purple (segmental
aneuploidy); group labels appear to the right:

```{r}
samples_non_euploid <- c(types_df$aneuploid, types_df$segmental)
if (length(samples_non_euploid) > 0) {
  p_summ <- compare_cn_track_summary(multi_esti,
                                     samples_to_plot = samples_non_euploid)
  p_summ
} else {
  message("No aneuploid or segmental samples found in multi_esti.")
}

samples_non_euploid_filtered <- c(types_df_filtered$aneuploid, types_df_filtered$segmental)
if (length(samples_non_euploid_filtered) > 0) {
  p_summ <- compare_cn_track_summary(multi_esti_filtered,
                                     samples_to_plot = samples_non_euploid_filtered)
  p_summ
} else {
  message("No aneuploid or segmental samples found in multi_esti_filtered.")
}
```

You can plot specific samples of interest by providing a vector of sample IDs to the `samples_to_plot` argument.

# Re-standardization (Optional)

If you want to improve BAF resolution after identifying a subset of samples with known ploidy via the
HMM, you can re-standardize using the estimated copy numbers as a reference.

Re-standardization requires access to the original raw data objects (`data` and `geno.pos`). Load them
before running this section if they are not already in your environment.

```{r}
# Recover original input columns; Chr -> Chromosome to match geno.pos schema
data <- qploidy_standardization$data[, c("MarkerName", "SampleName", "X", "Y", "R", "ratio")]

geno.pos <- unique(qploidy_standardization$data[, c("MarkerName", "Chr", "Position")])
colnames(geno.pos)[colnames(geno.pos) == "Chr"] <- "Chromosome"

# Depth thresholds used during re-standardization
min_depth <- quantile(data$R, 0.1, na.rm = TRUE)
max_depth <- quantile(data$R, 0.9, na.rm = TRUE)

re_qploidy_standardization <- re_standardize(
  data = data,
  geno.pos = geno.pos,
  hmm_CN_multi = multi_esti_filtered,
  use_estimated_dosages = TRUE,
  ploidy.standardization = ploidy_standardization,
  threshold.n.clusters = 4,
  threshold.missing.geno = 0.9,
  min.depth = min_depth,
  max.depth = max_depth,
  n.cores = n_cores,
  out_filename = file.path(output_dir, paste0(output_prefix, "_re_standardization.tsv.gz")),
  verbose = TRUE
)

re_qploidy_standardization
```

If `use_estimated_dosages = TRUE`, Qploidy2 estimates dosages from its own algorithm. If `FALSE`, it
uses the dosages in the initial `genos` object.

Compare BAF distributions before and after re-standardization for a sample:

```{r}
plot_qploidy_standardization(x = qploidy_standardization,
                             sample = sample_id,
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))
plot_qploidy_standardization(x = re_qploidy_standardization,
                             sample = sample_id,
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))
```

# Re-run the HMM with the New Standardization

```{r}
re_multi_esti <- hmm_estimate_CN_multi(
  qploidy_standarize_result = re_qploidy_standardization,
  sample_ids = "all",
  n_cores = n_cores,
  cn_grid = cn_grid)

re_multi_esti_filtered <- filter_hmm_CN(re_multi_esti,
                                        min_CN_reliability = min_CN_reliability,
                                        min_post_max = min_post_max)

count_types(re_multi_esti)


count_types(re_multi_esti_filtered)

# CNV profiles before re-standardization
compare_cn_track(hmm_CN = multi_esti_filtered,
                 samples_to_plot = samples[1:min(50, length(samples))],
                 facet_ncol = 8)

# CNV profiles after re-standardization
compare_cn_track(hmm_CN = re_multi_esti_filtered,
                 samples_to_plot = samples[1:min(50, length(samples))],
                 facet_ncol = 8)
```

# Refine Results for a Single Sample

Inspect a sample of interest before adjusting parameters. Choose from the aneuploid or segmental
samples identified above, or any sample that looks ambiguous:

```{r}
plot_karyotype(re_multi_esti_filtered, sample_name = sample_id, nrow = 2, color_by = "post_max_CN")

p <- plot_cn_track(hmm_CN = re_multi_esti_filtered,
                   qploidy_standarize_result = re_qploidy_standardization,
                   sample_id = sample_id,
                   show_window_lines = TRUE)
p$arranged
```

If the result is ambiguous, try increasing the number of markers per window to better fit the
mixture models on the BAF distributions:

```{r}
res_tweak <- hmm_estimate_CN(qploidy_standarize_result = re_qploidy_standardization,
                             sample_id = sample_id,
                             min_snps_per_window = 40,
                             cn_grid = cn_grid)

plot_karyotype(res_tweak, sample_name = sample_id, nrow = 2, color_by = "post_max_CN")

p <- plot_cn_track(hmm_CN = res_tweak,
                   qploidy_standarize_result = re_qploidy_standardization,
                   sample_id = sample_id,
                   show_window_lines = TRUE)
p$arranged
```

Explore more parameters with `?hmm_estimate_CN`.

Once you reach a conclusion, update the multi-sample HMM object:

```{r}
final_res <- update_hmm_CN_multi(re_multi_esti_filtered,
                                 res_tweak,
                                 rm_sample = FALSE)
```

Set `rm_sample = TRUE` to remove an inconclusive sample from the final results instead.

# Plotting All Samples in Batches (Optional)

Generate CNV plots for all samples in batches and save to disk:

```{r}
library(ggplot2)
n_plots_by_page <- c(seq(1, length(samples), by = 40), length(samples))
for (i in 1:(length(n_plots_by_page) - 1)) {
  comp_p <- compare_cn_track(hmm_CN = final_res,
                             samples_to_plot = samples[n_plots_by_page[[i]]:n_plots_by_page[i + 1]],
                             facet_ncol = 12)
  ggsave(comp_p,
         filename = file.path(images_dir, paste0(output_prefix, "_Samples_", n_plots_by_page[i], "_", n_plots_by_page[i + 1], ".png")),
         height = 10, width = 11)
}
```

# Interactive Curation of Results

**Warning**: Qploidy2 HMM accuracy depends on many factors that the call likelihoods may not always
capture. We strongly suggest reviewing diagnostic plots for unexpected results. The `GenoBrew` tool
described below provides a streamlined interface for visual evaluation and results curation.

## Export HMM Results

All objects produced by `update_hmm_CN_multi`, `hmm_estimate_CN`, and `hmm_estimate_CN_multi` share
the same class and can be exported with:

```{r}
write_hmm_CN(multi_esti, prefix = file.path(output_dir, paste0(output_prefix, "firstrun")))
write_hmm_CN(multi_esti_filtered, prefix = file.path(output_dir, paste0(output_prefix, "firstrun_filtered")))
write_hmm_CN(re_multi_esti, prefix = file.path(output_dir, paste0(output_prefix, "rerun")))
write_hmm_CN(re_multi_esti_filtered, prefix = file.path(output_dir, paste0(output_prefix, "rerun_filtered")))
```

This generates three files:

- `"[prefix]_by_marker.csv.gz"`
- `"[prefix]_by_window.csv.gz"`
- `"[prefix]_params.rds"`

Reload them in a new session with:

```{r, eval=FALSE}
final_res <- read_hmm_CN(
  by_marker_file = file.path(output_dir, paste0(output_prefix, "firstrun_by_marker.csv.gz")),
  by_window_file = file.path(output_dir, paste0(output_prefix, "firstrun_by_window.csv.gz")),
  params_file    = file.path(output_dir, paste0(output_prefix, "firstrun_params.rds"))
)
```

## Deploy GenoBrew

The files produced by `write_hmm_CN` can be uploaded to the `GenoBrew` Shiny interface. The
`CNV profiles` tab allows interactive inspection of BAF, Z-score and CN call plots, HMM re-running
with adjusted parameters for specific samples, and results updating.

Visit the `GenoBrew` GitHub [repository](https://github.com/Breeding-Insight/GenoBrew)
and [tutorial](https://scribehow.com/viewer/GenoBrew_Interactive_Marker_Panel_Evaluation_CNV_Visualization_and_Curation__4uWloBuPT1WlnCvW2UWiTg)
for more information.

# Call Dosages

**Warning**: Beta version — dosage call requires more testing and review.

Qploidy2 uses BAF distributions to call dosages.

Call dosages for all samples considering the copy number estimated:

```{r}
dosages <- call_hmm_dosages(hmm_CN = final_res, n.cores = n_cores)
head(dosages)
```

Visualize karyotype colored by dosage:

```{r}
plot_karyotype(dosages, sample_name = sample_id, nrow = 2, color_by = "dosage")
plot_karyotype(dosages, sample_name = sample_id, nrow = 2, color_by = "post_max_dosage")
```

# Export Results

## Summarized by Sample, Chromosome, Window, or Marker

```{r}
# By chromosome
summ_hmm_chr <- summarize_cn_mode(final_res, level = "chromosome")
head(summ_hmm_chr)

# By sample
summ_hmm_sample <- summarize_cn_mode(final_res, level = "sample")
head(summ_hmm_sample)

# By window
head(final_res$by_window)

# By marker
head(final_res$by_marker)
```

Export to CSV:

```{r, eval=FALSE}
write.csv(summ_hmm_chr,          file = file.path(output_dir, paste0(output_prefix, "_CN_by_chromosome.csv")))
write.csv(summ_hmm_sample,       file = file.path(output_dir, paste0(output_prefix, "_CN_by_sample.csv")))
write.csv(final_res$by_window,   file = file.path(output_dir, paste0(output_prefix, "_CN_by_window.csv")))
write.csv(final_res$by_marker,   file = file.path(output_dir, paste0(output_prefix, "_CN_by_marker.csv")))
```

## Export to MAPpoly filter_aneuploidy Function

`export_mappoly()` converts per-window copy-number calls into the input format expected by the
`filter_aneuploidy()` function in the [MAPpoly](https://github.com/mmollina/MAPpoly) package.
It returns a data frame with one row per sample, indicating the estimated CN for each chromosome:

```{r}
for_mappoly <- export_mappoly(final_res)
head(for_mappoly)
```

## Karyotype Notation

`karyotype_notation()` summarises per-marker copy-number calls into a compact table:

- **Ploidy**: overall ploidy
- **ChrID**: chromosome identifier
- **ChrCN**: most common CN for the chromosome
- **Segment**: position of segments with CN different from overall ploidy (ISCN-inspired notation:
  `g.start_end×CN`); `NA` if no CNV segments were detected

```{r}
k_not <- karyotype_notation(dosages)
head(k_not)

# Example: inspect a specific sample
k_not[which(k_not$SampleName == sample_id), ]
```

## VCF Format

```{r}
export_VCF(dosages, file = file.path(output_dir, paste0(output_prefix, "_CN_dosages.vcf")))
```

Read the VCF back with vcfR and extract fields:

```{r}
library(vcfR)
vcf <- read.vcfR(file.path(output_dir, paste0(output_prefix, "_CN_dosages.vcf")))
vcf@gt[1:5, 1:5]

CN <- extract.gt(vcf, "CN")
CN[1:5, 1:5]
```

The VCF dosages are adapted to the CN estimated by Qploidy2. For example, if CN = 3, dosages can be
0, 1, 2, 3 and are represented as `0/0/0`, `0/0/1`, `0/1/1`, `1/1/1`. If CN = 4, dosages range 0–4
and are represented as `0/0/0/0` through `1/1/1/1`.

```{r}
tri <- which(CN == "3", arr.ind = TRUE)
CN[tri[1:min(5, nrow(tri)), 1], 1:min(6, ncol(CN))]

GT <- extract.gt(vcf, "GT")
GT[tri[1:min(5, nrow(tri)), 1], 1:min(6, ncol(GT))]
```

The VCF contains the likelihood of each dosage call in the `PMD` field:

```{r}
dosage_prob <- extract.gt(vcf, "PMD")
dosage_prob[1:5, 1:5]
dosage_prob[tri[1:min(5, nrow(tri)), 1], 1:min(6, ncol(dosage_prob))]
```

# How to Cite

### Qploidy — Standardization

Taniguti, C. H., Lau, J., Hochhaus, T., Arias, D. C. L., Hokanson, S. C., Zlesak, D. C., Byrne, D. H.,
Klein, P. E., & Riera-Lizarazu, O. (2025). Exploring chromosomal variations in garden roses: Insights
from high-density SNP array data and a new tool, Qploidy. The Plant Genome, e70044.
https://doi.org/10.1002/tpg2.70044

### Qploidy2 — HMM and Grid Approach BAF Model Selection

Manuscript in preparation. Please contact the author for more information.

### nQuack — EM Approach BAF Model Selection

Gaynor, M., Landis, J., O'Connor, T., Laport, R., Doyle, J., Soltis, D., Ponciano, J., & Soltis, P.
(2024). "nQuack: An R package for predicting ploidal level from sequence data using site-based
heterozygosity." Applications in Plant Sciences, 12(4), e11606. doi:10.1002/aps3.11606

# Acknowledgments

### Qploidy version 1.0.0

Funded in part by the Robert E. Basye Endowment in Rose Genetics, Dept. of Horticultural Sciences,
Texas A&M University, and USDA's National Institute of Food and Agriculture (NIFA), Specialty Crop
Research Initiative (SCRI) projects: "Tools for Genomics-Assisted Breeding in Polyploids: Development
of a Community Resource" (Award No. 2020-51181-32156); and "Developing Sustainable Rose Landscapes via
Rose Rosette Disease Education, Socioeconomic Assessments, and Breeding RRD-Resistant Roses with Stable
Black Spot Resistance" (Award No. 2022-51181-38330).

### Qploidy versions > 1.0.0 and Qploidy2

Supported by [Breeding Insight](https://www.breedinginsight.org/).
