---
title: "Qploidy2 QuickStart Template"
subtitle: "Copy number estimation from standardized data"
author: "Breeding Insight"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    css: BIcolors_template.css
    includes:
      in_header: header.html
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 10,
  out.width = "40%"
)
```

# Parameters

Define key parameters before running the analysis. Adjust these to match your dataset.

```{r parameters}
# Path to your standardized data file
standardization_file <- "standardization.tsv.gz"

# Index of the example sample used throughout this template for single-sample steps.
# Increase if samples[3] is not informative (e.g., low heterozygosity).
example_sample_idx <- 3

# Number of CPU cores for parallel computation
n_cores <- 2

# Copy number range to test (adjust based on your expected ploidy)
cn_grid <- c(2, 3, 4, 5, 6)

# BAF kernel bandwidth values to evaluate
bw <- c(0.02, 0.04, 0.06, 0.08, 0.1)

# Predominant ploidy for re-standardization (should match the most frequent ploidy in your dataset)
ploidy_standardization <- 4

# Depth thresholds used during re-standardization
min_depth <- 30
max_depth <- 600

# Quality thresholds for filter_hmm_CN
min_CN_reliability <- 0.25
min_post_max <- 0.6

# Output file prefix for exported results
output_prefix <- "qploidy_results"
```

# Introduction

This template demonstrates the Qploidy2 workflow starting from a pre-computed `standardization.tsv.gz`
file. It covers quality assessment, BAF model selection, Hidden Markov Model (HMM)-based copy number
estimation, visualization, interactive curation, and VCF export.

If you have not yet produced the standardization file, follow the full tutorial
(`Qploidy_alfalfa_tutorial.Rmd`) for input preparation and standardization steps.

# Load Standardized Data

```{r}
library(Qploidy2)

qploidy_standardization <- read_qploidy_standardization(standardization_file)
samples <- unique(qploidy_standardization$data$SampleName)
sample_id <- samples[example_sample_idx]

cat("Loaded", length(samples), "samples\n")
cat("Example sample:", sample_id, "\n")
```

# Quality Assessment and Visualization

Visualize the effect of standardization for a single sample and explore BAF and Z-score distributions:

```{r}
plot_qploidy_standardization(x = qploidy_standardization,
                             sample = sample_id,
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))

plot_qploidy_standardization(x = qploidy_standardization,
                             sample = sample_id,
                             type = c("zscore", "R"))
```

See `?plot_qploidy_standardization` for all available plot types.

## Chromosome-level screening for abnormalities based on Z-score (optional)

Standardized Z-scores can be used to rapidly identify candidate samples with potential aneuploidies
or large CNVs before running the full HMM, and to select samples with known ploidy to guide BAF model
selection.

```{r}
chromosome_level_test_plot_qploidy(input = qploidy_standardization,
                                   selected_samples = samples[1:12],
                                   col2test = "z")
```

# BAF Model Selection

Select the best BAF model for a sample. The approach implemented here tests different statistical
distribution mixture models for each candidate ploidy on the BAF distributions and selects the best
fit based on BIC.

There are two approaches available:

- **Grid approach**: Tests a grid of parameters — less computationally demanding but may not provide
  the most tailored parameters. The results from this step serve as initial parameters for the HMM,
  which will be further optimized by the EM algorithm implemented in the HMM.
- **EM approach**: Uses a robust EM algorithm to return optimized parameters. See tutorial
  *Qploidy2 and nQuack integration* for details.

This template applies the grid approach. You can select a sample with known ploidy (e.g., identified
in the chromosome-level screening above) to guide model selection.

```{r}
selected_model <- select_best_baf_model(qploidy_standardization = qploidy_standardization,
                                        sample = sample_id,
                                        cn_grid = cn_grid,
                                        bw = bw,
                                        plot = TRUE)

selected_model$plot
head(selected_model$grid_results)
```

# HMM Copy Number Estimation (Single Sample)

Estimate copy number for a sample using a Hidden Markov Model. This step integrates standardized BAF
and Z-score information across genomic regions to infer CNV segments.
Find all parameter descriptions in `?hmm_estimate_CN`.

**Warning**: Avoid testing copy numbers that are not biologically plausible for your species to reduce
the number of evaluated scenarios and improve runtime.

```{r}
res <- hmm_estimate_CN(qploidy_standarize_result = qploidy_standardization,
                       sample_id = sample_id,
                       cn_grid = cn_grid)

res
```

Note that the HMM multipoint approach considers marker position order based on the reference genome
used. Therefore, changing the reference genome may change the results.

`Qploidy2` offers several ways to visualize the estimated karyotype:

```{r}
plot_karyotype(res, sample_name = sample_id, nrow = 2)
```

Color by the CN call final likelihood (how confident the model was for each marker):

```{r}
plot_karyotype(res, sample_name = sample_id, nrow = 2, color_by = "post_max_CN")
```

Color by the BAF weight — windows inferred from both BAF and Z-score have weight 0.5; windows with no
heterozygous genotypes are inferred from Z-score only and have BAF weight close to 0:

```{r}
plot_karyotype(res, sample_name = sample_id, nrow = 2, color_by = "w_baf")
```

The `plot_cn_track` function provides a detailed diagnostic of the process:

```{r}
p <- plot_cn_track(hmm_CN = res,
                   qploidy_standarize_result = qploidy_standardization,
                   sample_id = sample_id,
                   show_window_lines = FALSE,
                   z_by_mean = TRUE)

p$arranged
```

The CN segments in the bottom track are coloured by `CN_reliability`, a joint confidence score
calculated as `post_max × normalized_w_baf`. It summarises both the HMM's certainty about the
called CN state and whether the call was supported by BAF evidence:

- **Range 0–1**: a score of `1` means the HMM assigned the highest posterior to the called CN
  state *and* the window contains enough heterozygous markers for BAF support to also be
  confident.
- **Low values near 0**: the call relies almost entirely on Z-score (total depth) because the
  window lacks heterozygous markers (`w_baf ≈ 0`). By default, such windows are capped at
  `0.25 × post_max`, so even a perfect HMM call cannot exceed 0.25 without BAF evidence.
- **Interpreting the plot**: bright segments indicate high-confidence calls backed by both
  signals; faded segments warrant manual inspection. Use `filter_hmm_CN(min_CN_reliability = ...)`
  in the [Filter by CN Call Quality Metrics](#filter-by-cn-call-quality-metrics) section to
  mask low-confidence windows before export.

```{r}
# Individual panels
p$baf
p$z
p$cn
p$baf_dist
```

View results in `data.frame` format:

```{r}
head(res$by_window)
head(res$by_marker)
head(res$params)
```

# HMM Copy Number Estimation (Multiple Samples)

Run HMM for all samples in parallel. The BAF model is selected internally using
`select_best_baf_model`; define the grids (`bw`, `add_uniform_grid`, and `uniform_weight_grid`)
according to what you observed in the single-sample step above.

**Note**: Running HMM for many samples can be computationally intensive. Adjust `n_cores` based on
your available resources and dataset size.

```{r}
multi_esti <- hmm_estimate_CN_multi(qploidy_standarize_result = qploidy_standardization,
                                    sample_ids = "all",
                                    n_cores = n_cores,
                                    cn_grid = cn_grid,
                                    bw = bw)
```

Visualize a CNV summary for multiple samples:

```{r}
compare_cn_track(hmm_CN = multi_esti,
                 samples_to_plot = samples[1:min(30, length(samples))],
                 facet_ncol = 8)
```

# Filter by CN Call Quality Metrics

`filter_hmm_CN()` masks CN calls that do not meet user-defined quality thresholds by replacing
`CN_call`, `post_max`, `CN_reliability`, and all `post_CN*` posteriors with `NA`. Masked windows are
treated as missing by all downstream plotting and export functions.

The two most commonly used thresholds are:

* `min_CN_reliability` — joint confidence combining HMM posterior probability and BAF weight.
  A value of `0` means no BAF evidence at all; `1` means full BAF support with a confident HMM call.
  Windows supported only by Z-score cannot exceed `z_no_baf_scale × post_max` in reliability
  (default cap = `0.25 × post_max`).
* `min_post_max` — the raw HMM posterior probability for the called CN state, regardless of BAF
  support. Useful when you want to retain Z-score-only calls but still discard low-confidence HMM
  calls.

```{r}
multi_esti_filtered <- filter_hmm_CN(multi_esti,
                                     min_CN_reliability = min_CN_reliability,
                                     min_post_max = min_post_max)
```

# Summarized Plots and Tables

After filtering, get a high-level overview of the entire population.

`count_types()` classifies every sample as **euploid** (all chromosomes match the sample-level copy
number), **aneuploid** (one or more chromosomes carry a different but uniform CN — whole-chromosome
gain or loss), or **segmental** (at least one chromosome has windows with mixed CN, indicating a
partial chromosomal change):

```{r}
types_df <- count_types(multi_esti)
types_df
str(types_df)

types_df_filtered <- count_types(multi_esti_filtered)
types_df_filtered
str(types_df_filtered)
```

Use the returned lists to visualize samples of each type:

```{r}
# Adjust indices if the first sample of a type is not informative
plot_karyotype(multi_esti,
               sample_name = types_df$aneuploid[1], color_by = "CN_reliability")

# Option to remove ghost copies:
plot_karyotype(multi_esti,
               sample_name = types_df$aneuploid[1], color_by = "CN_reliability", add_ghost = FALSE)

plot_karyotype(multi_esti_filtered,
               sample_name = types_df_filtered$segmental[1], color_by = "CN_reliability")
```

`compare_cn_track_summary()` produces a karyotype-style matrix with samples in rows and chromosomes
in columns. Samples are ordered by ascending sample-level CN and, within each CN block, by ploidy
category (euploid → aneuploid → segmental). Chromosomes with mixed CN are shown in purple (segmental
aneuploidy); group labels appear to the right:

```{r}
p_summ <- compare_cn_track_summary(multi_esti,
                                   samples_to_plot = c(types_df$aneuploid, types_df$segmental))
p_summ

p_summ <- compare_cn_track_summary(multi_esti_filtered,
                                   samples_to_plot = c(types_df_filtered$aneuploid,
                                                       types_df_filtered$segmental))
p_summ
```

You can plot specific samples of interest by providing a vector of sample IDs to the `samples_to_plot` argument.

# Re-standardization (Optional)

If you want to improve BAF resolution after identifying a subset of samples with known ploidy via the
HMM, you can re-standardize using the estimated copy numbers as a reference.

Re-standardization requires access to the original raw data objects (`data` and `geno.pos`). Load them
before running this section if they are not already in your environment.

```{r}
re_qploidy_standardization <- re_standardize(
  data = data,
  geno.pos = geno.pos,
  hmm_CN_multi = multi_esti_filtered,
  use_estimated_dosages = TRUE,
  ploidy.standardization = ploidy_standardization,
  threshold.n.clusters = 4,
  threshold.missing.geno = 0.9,
  min.depth = min_depth,
  max.depth = max_depth,
  n.cores = n_cores,
  out_filename = "re_standardization.tsv.gz",
  verbose = TRUE
)

re_qploidy_standardization
```

If `use_estimated_dosages = TRUE`, Qploidy2 estimates dosages from its own algorithm. If `FALSE`, it
uses the dosages in the initial `genos` object.

Compare BAF distributions before and after re-standardization for a sample:

```{r}
plot_qploidy_standardization(x = qploidy_standardization,
                             sample = sample_id,
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))
plot_qploidy_standardization(x = re_qploidy_standardization,
                             sample = sample_id,
                             type = c("Ratio_hist_overall", "BAF_hist_overall"))
```

# Re-run the HMM with the New Standardization

```{r}
re_multi_esti <- hmm_estimate_CN_multi(
  qploidy_standarize_result = re_qploidy_standardization,
  sample_ids = "all",
  n_cores = n_cores,
  cn_grid = cn_grid,
  bw = bw)

re_multi_esti_filtered <- filter_hmm_CN(re_multi_esti,
                                        min_CN_reliability = min_CN_reliability,
                                        min_post_max = min_post_max)

# CNV profiles before re-standardization
compare_cn_track(hmm_CN = multi_esti_filtered,
                 samples_to_plot = samples[1:min(50, length(samples))],
                 facet_ncol = 8)

# CNV profiles after re-standardization
compare_cn_track(hmm_CN = re_multi_esti_filtered,
                 samples_to_plot = samples[1:min(50, length(samples))],
                 facet_ncol = 8)
```

# Refine Results for a Single Sample

Inspect a sample of interest before adjusting parameters. Choose from the aneuploid or segmental
samples identified above, or any sample that looks ambiguous:

```{r}
plot_karyotype(re_multi_esti_filtered, sample_name = sample_id, nrow = 2, color_by = "post_max_CN")

p <- plot_cn_track(hmm_CN = re_multi_esti_filtered,
                   qploidy_standarize_result = re_qploidy_standardization,
                   sample_id = sample_id,
                   show_window_lines = TRUE)
p$arranged
```

If the result is ambiguous, try increasing the number of markers per window to better fit the
mixture models on the BAF distributions:

```{r}
res_tweak <- hmm_estimate_CN(qploidy_standarize_result = re_qploidy_standardization,
                             sample_id = sample_id,
                             min_snps_per_window = 40,
                             cn_grid = cn_grid)

plot_karyotype(res_tweak, sample_name = sample_id, nrow = 2, color_by = "post_max_CN")

p <- plot_cn_track(hmm_CN = res_tweak,
                   qploidy_standarize_result = re_qploidy_standardization,
                   sample_id = sample_id,
                   show_window_lines = TRUE)
p$arranged
```

Explore more parameters with `?hmm_estimate_CN`.

Once you reach a conclusion, update the multi-sample HMM object:

```{r}
final_res <- update_hmm_CN_multi(re_multi_esti_filtered,
                                 res_tweak,
                                 rm_sample = FALSE)
```

Set `rm_sample = TRUE` to remove an inconclusive sample from the final results instead.

# Plotting All Samples in Batches (Optional)

Generate CNV plots for all samples in batches and save to disk:

```{r, eval=FALSE}
n_plots_by_page <- c(seq(1, length(samples), by = 40), length(samples))
for (i in 1:(length(n_plots_by_page) - 1)) {
  comp_p <- compare_cn_track(hmm_CN = final_res,
                             samples_to_plot = samples[n_plots_by_page[[i]]:n_plots_by_page[i + 1]],
                             facet_ncol = 12)
  ggsave(comp_p,
         filename = paste0("Samples_", n_plots_by_page[i], "_", n_plots_by_page[i + 1], ".png"),
         height = 10, width = 11)
}
```

# Interactive Curation of Results

**Warning**: Qploidy2 HMM accuracy depends on many factors that the call likelihoods may not always
capture. We strongly suggest reviewing diagnostic plots for unexpected results. The `GenoBrew` tool
described below provides a streamlined interface for visual evaluation and results curation.

## Export HMM Results

All objects produced by `update_hmm_CN_multi`, `hmm_estimate_CN`, and `hmm_estimate_CN_multi` share
the same class and can be exported with:

```{r}
write_hmm_CN(final_res, prefix = output_prefix)
```

This generates three files:

- `"[prefix]_by_marker.csv.gz"`
- `"[prefix]_by_window.csv.gz"`
- `"[prefix]_params.rds"`

Reload them in a new session with:

```{r, eval=FALSE}
final_res <- read_hmm_CN(
  by_marker_file = paste0(output_prefix, "_by_marker.csv.gz"),
  by_window_file = paste0(output_prefix, "_by_window.csv.gz"),
  params_file    = paste0(output_prefix, "_params.rds")
)
```

## Deploy GenoBrew

The files produced by `write_hmm_CN` can be uploaded to the `GenoBrew` Shiny interface. The
`CNV profiles` tab allows interactive inspection of BAF, Z-score and CN call plots, HMM re-running
with adjusted parameters for specific samples, and results updating.

Visit the `GenoBrew` GitHub [repository](https://github.com/Breeding-Insight/GenoBrew)
and [tutorial](https://scribehow.com/viewer/GenoBrew_Interactive_Marker_Panel_Evaluation_CNV_Visualization_and_Curation__4uWloBuPT1WlnCvW2UWiTg)
for more information.

# Call Dosages

**Warning**: Beta version — dosage call requires more testing and review.

Qploidy2 uses BAF distributions to call dosages. Example for a single sample:

```{r}
baf <- final_res$by_marker$baf[final_res$by_marker$SampleName == sample_id]

selected_model_dosage <- select_best_baf_model(baf_vec = baf,
                                               sample = sample_id,
                                               cn_grid = cn_grid,
                                               bw = bw,
                                               plot = TRUE)

dosages_one_sample <- call_BAF_dosages(baf, selected_model = selected_model_dosage, plot = TRUE)
head(dosages_one_sample$data)

dosages_one_sample$plot
```

Call dosages for all samples considering the copy number estimated:

```{r}
dosages <- call_hmm_dosages(hmm_CN = final_res, n.cores = n_cores)
head(dosages)
```

Visualize karyotype colored by dosage:

```{r}
plot_karyotype(dosages, sample_name = sample_id, nrow = 2, color_by = "dosage")
plot_karyotype(dosages, sample_name = sample_id, nrow = 2, color_by = "post_max_dosage")
```

# Export Results

## Summarized by Sample, Chromosome, Window, or Marker

```{r}
# By chromosome
summ_hmm_chr <- summarize_cn_mode(final_res, level = "chromosome")
head(summ_hmm_chr)

# By sample
summ_hmm_sample <- summarize_cn_mode(final_res, level = "sample")
head(summ_hmm_sample)

# By window
head(final_res$by_window)

# By marker
head(final_res$by_marker)
```

Export to CSV:

```{r, eval=FALSE}
write.csv(summ_hmm_chr,          file = paste0(output_prefix, "_CN_by_chromosome.csv"))
write.csv(summ_hmm_sample,       file = paste0(output_prefix, "_CN_by_sample.csv"))
write.csv(final_res$by_window,   file = paste0(output_prefix, "_CN_by_window.csv"))
write.csv(final_res$by_marker,   file = paste0(output_prefix, "_CN_by_marker.csv"))
```

## Export to MAPpoly filter_aneuploidy Function

`export_mappoly()` converts per-window copy-number calls into the input format expected by the
`filter_aneuploidy()` function in the [MAPpoly](https://github.com/mmollina/MAPpoly) package.
It returns a data frame with one row per sample, indicating the estimated CN for each chromosome:

```{r}
for_mappoly <- export_mappoly(final_res)
head(for_mappoly)
```

## Karyotype Notation

`karyotype_notation()` summarises per-marker copy-number calls into a compact table:

- **Ploidy**: overall ploidy
- **ChrID**: chromosome identifier
- **ChrCN**: most common CN for the chromosome
- **Segment**: position of segments with CN different from overall ploidy (ISCN-inspired notation:
  `g.start_end×CN`); `NA` if no CNV segments were detected

```{r}
k_not <- karyotype_notation(dosages)
head(k_not)

# Example: inspect a specific sample
k_not[which(k_not$SampleName == sample_id), ]
```

## VCF Format

```{r}
export_VCF(dosages, file = paste0(output_prefix, "_CN_dosages.vcf"))
```

Read the VCF back with vcfR and extract fields:

```{r}
library(vcfR)
vcf <- read.vcfR(paste0(output_prefix, "_CN_dosages.vcf"))
vcf@gt[1:5, 1:5]

CN <- extract.gt(vcf, "CN")
CN[1:5, 1:5]
```

The VCF dosages are adapted to the CN estimated by Qploidy2. For example, if CN = 3, dosages can be
0, 1, 2, 3 and are represented as `0/0/0`, `0/0/1`, `0/1/1`, `1/1/1`. If CN = 4, dosages range 0–4
and are represented as `0/0/0/0` through `1/1/1/1`.

```{r}
tri <- which(CN == "3", arr.ind = TRUE)
CN[tri[1:min(5, nrow(tri)), 1], 1:min(6, ncol(CN))]

GT <- extract.gt(vcf, "GT")
GT[tri[1:min(5, nrow(tri)), 1], 1:min(6, ncol(GT))]
```

The VCF contains the likelihood of each dosage call in the `PMD` field:

```{r}
dosage_prob <- extract.gt(vcf, "PMD")
dosage_prob[1:5, 1:5]
dosage_prob[tri[1:min(5, nrow(tri)), 1], 1:min(6, ncol(dosage_prob))]
```

# How to Cite

### Qploidy — Standardization

Taniguti, C. H., Lau, J., Hochhaus, T., Arias, D. C. L., Hokanson, S. C., Zlesak, D. C., Byrne, D. H.,
Klein, P. E., & Riera-Lizarazu, O. (2025). Exploring chromosomal variations in garden roses: Insights
from high-density SNP array data and a new tool, Qploidy. The Plant Genome, e70044.
https://doi.org/10.1002/tpg2.70044

### Qploidy2 — HMM and Grid Approach BAF Model Selection

Manuscript in preparation. Please contact the author for more information.

### nQuack — EM Approach BAF Model Selection

Gaynor, M., Landis, J., O'Connor, T., Laport, R., Doyle, J., Soltis, D., Ponciano, J., & Soltis, P.
(2024). "nQuack: An R package for predicting ploidal level from sequence data using site-based
heterozygosity." Applications in Plant Sciences, 12(4), e11606. doi:10.1002/aps3.11606

# Acknowledgments

### Qploidy version 1.0.0

Funded in part by the Robert E. Basye Endowment in Rose Genetics, Dept. of Horticultural Sciences,
Texas A&M University, and USDA's National Institute of Food and Agriculture (NIFA), Specialty Crop
Research Initiative (SCRI) projects: "Tools for Genomics-Assisted Breeding in Polyploids: Development
of a Community Resource" (Award No. 2020-51181-32156); and "Developing Sustainable Rose Landscapes via
Rose Rosette Disease Education, Socioeconomic Assessments, and Breeding RRD-Resistant Roses with Stable
Black Spot Resistance" (Award No. 2022-51181-38330).

### Qploidy versions > 1.0.0 and Qploidy2

Supported by [Breeding Insight](https://www.breedinginsight.org/).
